Splitting it up: the \pkg{spduration} split-population duration regression package

### 2016-02-19 manuscript.Rmd
style and bibliography changes for JSS

---
author:
  - name: Andreas Beger
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: ab428@duke.edu
    url: something
  - name: Daniel W. Hill, Jr.
    affiliation: University of Georgia
    address: >
      Department of International Affairs, University of Georgia
      Athens, GA, USA
    email: dwhill@uga.edu
  - name: Nils W. Metternich
    affiliation: University College London
    address: >
      Department of Political Science, University College London
      London, UK
    email: n.metternich@ucl.ac.uk
  - name: Shahryar Minhas
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: sfm12@duke.edu
  - name: Michael D. Ward
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: mw160@duke.edu
title:
  formatted: "Splitting it up: the \\pkg{spduration} split-population duration regression package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration models. The package includes estimation and several post-estimation methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [regression, rare events, "\\proglang{R}"]
  plain:     [regression, rare events, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

However, there are conditions under which not all individuals have the same underlying risk to experience a specific outcome and might not even be at risk at all. Consider the risk of acquiring a viral infection like the common flu. Let us initially assume that everyone is at risk of infection, and individual behavior (e.g. good hygiene) and structural factors (e.g. workplace) determine whether individuals catch the flu. Under these assumptions a standard duration model should provide us with efficient estimates of how different behavioral and structural factors impact on the general baseline risk. But, there might be individuals that are immune to the flu in the overall population--because they received vaccination, had the specific flu virus in the past, or have some other characteristic that makes it impossible for them to attract the disease. In this instance we have two underlying populations: an at risk population and an immune one. If the immune population is relatively large, estimates using a standard duration model will be biased and predictions inaccurate.

@@

The major modeling question in this setting, which we will return to below, is the choice of a function form (e.g. exponential, Weibull, or log-logistic) describing the underlying hazard rate $h(t) = \frac{f(t)}{S(t)}$ over time.

@@

Where $z_i$ is a vector of covariates for a subject at a given time. For interpretation, it is important to note that with time-varying covariates, the risk (or cured) estimate for a subject is particular to a given time point rather than constant over all time periods in the spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time. Subjects can theoretically have multiple spells, e.g. cancer patients who go into remission and relapse more than once, or states that experience multiple civil wars over their history.} Depending on the covariates, the risk estimate for a subject can thus fluctuate over time. To ease interpretation, it might be convenient to restrict covariates in the logit risk model to slow-moving, stable covariates in order to produce stable risk estimates for subjects.

@@

```{r data, warning=FALSE}
library("spduration")
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                            freq="year", ongoing = FALSE)
```

@@

\begin{table}
\begin{center}
\caption{Example data frame with select duration variables}
\label{tab-ex}
\begin{tabular}{lrrrrr} \toprule
Country & Year & Coup & Spell ID & Duration \\ \midrule
Portugal & 1960 & 0 & 1 & 1  \\ 
\ldots & \ldots & 0 & 1 & \ldots \\
\ldots & 1975 & 1 & 1 & 16 \\ \midrule
Portugal & 1976 & 0 & 2 & 1 \\
\ldots & \ldots & 0 & 2 & \ldots \\
\ldots & 2000 & 0 & 2 & 20  \\ \midrule
Canada & 1960 & 0 & 3 & 1 \\
\ldots & \ldots & 0 & 3 & \ldots \\ 
\ldots & 2000 & 0 & 3 & 41 \\ \bottomrule
\end{tabular}
\end{center}
\end{table}

@@

```{r models, include = TRUE, cache = TRUE}
weib_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data=bscoup, silent = TRUE)

loglog_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data=bscoup, distr="loglog", silent = TRUE)
```

@@

```{r summary-table, include = TRUE, eval = FALSE}
library("xtable")
tbl <- xtable(loglog_model, caption="Coup model with log-logistic hazard",
              label="loglog_table")
print(tbl, caption.placement="top", comment=FALSE, include.rownames=FALSE)
```

\begin{table}[ht]
\centering
\caption{Coup model with log-logistic hazard} 
\label{loglog_table}
\begin{tabular}{lrrr}
  \hline
Parameter & Estimate & StdErr & p \\ 
  \hline
(Dur. Intercept) & 2.40 & 0.21 & 0.00 \\ 
  milreg & -1.13 & 0.21 & 0.00 \\ 
  instab & -0.09 & 0.02 & 0.00 \\ 
  regconf & -2.52 & 2.16 & 0.24 \\ 
  log(alpha) & -0.45 & 0.06 & 0.00 \\ 
  (Risk Intercept) & 2.93 & 1.87 & 0.12 \\ 
  couprisk & 0.59 & 0.32 & 0.07 \\ 
  wealth & -0.36 & 0.28 & 0.20 \\ 
  milreg.1 & 10.82 & 9.31 & 0.25 \\ 
  rwar & -0.52 & 0.94 & 0.58 \\ 
  regconf.1 & -5.43 & 5.62 & 0.33 \\ 
  samerica & 2.09 & 1.45 & 0.15 \\ 
  camerica & -0.39 & 0.73 & 0.59 \\ 
   \hline
\end{tabular}
\end{table}

@@

```{r, eval = FALSE}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
                          freq="year", ongoing = FALSE)
```

```{r, include = FALSE}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup=="yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID="countryid", tID="year",
                          freq="year", ongoing = FALSE)
```

@@

```{r, include = FALSE}
coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                           freq="year", ongoing = FALSE)
coup_test  <- coup_test[coup_test$year >= 1996, ]
```

```{r, eval = FALSE}
coup_test  <- add_duration(bscoup, "coup", unitID="countryid", tID="year",
                           freq="year", ongoing = FALSE)
coup_test  <- coup_test[coup_test$year >= 1996, ]
```

@@

loglog_model2 <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = coup_train, distr="loglog", silent = TRUE) 
```

@@

par(mfrow=c(2,1),mar=c(2,2,2,2))
separationplot(weib2_test_p,   obs_y)
separationplot(loglog2_test_p, obs_y)
```

@@

Furthermore, as the example above shows, in out-of-sample testing based on some kind of data partitioning scheme, this coding scheme can lead to unintentional contamination of in-sample cases with knowledge of out-of-sample failures through the risk coding for failed spells. This leads to two recommendations. First, data should be partitioned spell or block-wise, e.g. by withholding the last *x* years of data, and not randomly. Second, given the two concerns of left-censoring and non-independence induced through the duration and risk coding, care should be taken to ensure that duration data are built without access to future information in another data partition.

>>>>

---
author:
  - name: Andreas Beger
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: ab428@duke.edu
    url: something
  - name: Daniel W. Hill, Jr.
    affiliation: University of Georgia
    address: >
      Department of International Affairs, University of Georgia
      Athens, GA, USA
    email: dwhill@uga.edu
  - name: Nils W. Metternich
    affiliation: University College London
    address: >
      Department of Political Science, University College London
      London, UK
    email: n.metternich@ucl.ac.uk
  - name: Shahryar Minhas
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: sfm12@duke.edu
  - name: Michael D. Ward
    affiliation: Duke University
    address: >
      Department of Political Science, Duke University
      Durham, NC, USA
    email: mw160@duke.edu
title:
  formatted: "Splitting it Up: The \\pkg{spduration} Split-population Duration Regression Package"
  # If you use tex in the formatted title, also supply version without
  plain:     "Splitting it up: the spduration split-population duration regression package"
  # For running headers, if needed
  short:     "\\pkg{spduration}: Split-population duration regression"
abstract: >
  We present an implementation of split-population duration regression in the   \pkg{spduration} R package and an application to data on military coups. The statistical model accounts for units that are immune to a certain outcome and not part of the duration process the researcher is primarily interested in. We provide insights that if immune units exist, we can significantly increase the predictive performance compared to standard duration models. The package includes estimation and several post-estimation methods for split-population Weibull and Loglogistic models.
keywords:
  # at least one keyword must be supplied
  formatted: [regression, rare events, "\\proglang{R}"]
  plain:     [regression, rare events, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{booktabs}
output: 
  rticles::jss_article
bibliography: spdur.bib
---

@@

However, there are conditions under which not all individuals have the same underlying risk to experience a specific outcome and might not even be at risk at all. Consider the risk of acquiring a viral infection like the common flu. Let us initially assume that everyone is at risk of infection, and individual behavior (e.g., good hygiene) and structural factors (e.g., workplace) determine whether individuals catch the flu. Under these assumptions a standard duration model should provide us with efficient estimates of how different behavioral and structural factors impact on the general baseline risk. But, there might be individuals that are immune to the flu in the overall population--because they received vaccination, had the specific flu virus in the past, or have some other characteristic that makes it impossible for them to attract the disease. In this instance we have two underlying populations: an at risk population and an immune one. If the immune population is relatively large, estimates using a standard duration model will be biased and predictions inaccurate.

@@

The major modeling question in this setting, which we will return to below, is the choice of a function form (e.g., exponential, Weibull, or log-logistic) describing the underlying hazard rate $h(t) = \frac{f(t)}{S(t)}$ over time.

@@

Where $z_i$ is a vector of covariates for a subject at a given time. For interpretation, it is important to note that with time-varying covariates, the risk (or cured) estimate for a subject is particular to a given time point rather than constant over all time periods in the spell.\footnote{We use ``spell'' to designate all time periods observed for a subject up to the failure time. Subjects can theoretically have multiple spells, e.g., cancer patients who go into remission and relapse more than once, or states that experience multiple civil wars over their history.} Depending on the covariates, the risk estimate for a subject can thus fluctuate over time. To ease interpretation, it might be convenient to restrict covariates in the logit risk model to slow-moving, stable covariates in order to produce stable risk estimates for subjects.

@@

```{r data, warning=FALSE}
library("spduration")
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
bscoup      <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
                            freq = "year", ongoing = FALSE)
```

@@

\begin{table}
\begin{center}
\begin{tabular}{lrrrrr} \toprule
Country & Year & Coup & Spell ID & Duration \\ \midrule
Portugal & 1960 & 0 & 1 & 1  \\ 
\ldots & \ldots & 0 & 1 & \ldots \\
\ldots & 1975 & 1 & 1 & 16 \\ \midrule
Portugal & 1976 & 0 & 2 & 1 \\
\ldots & \ldots & 0 & 2 & \ldots \\
\ldots & 2000 & 0 & 2 & 20  \\ \midrule
Canada & 1960 & 0 & 3 & 1 \\
\ldots & \ldots & 0 & 3 & \ldots \\ 
\ldots & 2000 & 0 & 3 & 41 \\ \bottomrule
\end{tabular}
\end{center}
\caption{Example data frame with select duration variables} \label{tab-ex}
\end{table}

@@

```{r models, include = TRUE, cache = TRUE}
weib_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = bscoup, silent = TRUE)

loglog_model <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = bscoup, distr = "loglog", silent = TRUE)
```

@@

```{r summary-table, include = TRUE, eval = FALSE}
library("xtable")
tbl <- xtable(loglog_model, caption = "Coup model with log-logistic hazard",
              label = "loglog_table")
print(tbl, caption.placement = "top", comment = FALSE, include.rownames = FALSE)
```

\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
Parameter & Estimate & StdErr & p \\ 
  \hline
(Dur. Intercept) & 2.40 & 0.21 & 0.00 \\ 
  milreg & -1.13 & 0.21 & 0.00 \\ 
  instab & -0.09 & 0.02 & 0.00 \\ 
  regconf & -2.52 & 2.16 & 0.24 \\ 
  log(alpha) & -0.45 & 0.06 & 0.00 \\ 
  (Risk Intercept) & 2.93 & 1.87 & 0.12 \\ 
  couprisk & 0.59 & 0.32 & 0.07 \\ 
  wealth & -0.36 & 0.28 & 0.20 \\ 
  milreg.1 & 10.82 & 9.31 & 0.25 \\ 
  rwar & -0.52 & 0.94 & 0.58 \\ 
  regconf.1 & -5.43 & 5.62 & 0.33 \\ 
  samerica & 2.09 & 1.45 & 0.15 \\ 
  camerica & -0.39 & 0.73 & 0.59 \\ 
   \hline
\end{tabular}
\caption{Coup model with log-logistic hazard} \label{loglog_table}
\end{table}

@@

```{r, eval = FALSE}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID = "countryid", tID = "year",
                          freq = "year", ongoing = FALSE)
```

```{r, include = FALSE}
data(bscoup)
bscoup$coup <- ifelse(bscoup$coup == "yes", 1, 0)
coup_train <- bscoup[bscoup$year < 1996, ]
coup_train <- add_duration(coup_train, "coup", unitID = "countryid", tID = "year",
                          freq = "year", ongoing = FALSE)
```

@@

```{r, include = FALSE}
coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
                           freq = "year", ongoing = FALSE)
coup_test  <- coup_test[coup_test$year >= 1996, ]
```

```{r, eval = FALSE}
coup_test  <- add_duration(bscoup, "coup", unitID = "countryid", tID = "year",
                           freq = "year", ongoing = FALSE)
coup_test  <- coup_test[coup_test$year >= 1996, ]
```

@@

loglog_model2 <- spdur(
  duration ~ milreg + instab + regconf,
  atrisk ~ couprisk + wealth + milreg + rwar + regconf + samerica + camerica,
  data = coup_train, distr = "loglog", silent = TRUE) 
```

@@

par(mfrow = c(2,1),mar = c(2,2,2,2))
separationplot(weib2_test_p,   obs_y)
separationplot(loglog2_test_p, obs_y)
```

@@

Furthermore, as the example above shows, in out-of-sample testing based on some kind of data partitioning scheme, this coding scheme can lead to unintentional contamination of in-sample cases with knowledge of out-of-sample failures through the risk coding for failed spells. This leads to two recommendations. First, data should be partitioned spell or block-wise, e.g., by withholding the last *x* years of data, and not randomly. Second, given the two concerns of left-censoring and non-independence induced through the duration and risk coding, care should be taken to ensure that duration data are built without access to future information in another data partition.
